---
title: "Complaints Against Police In Different Cities"
author: "Adina Scheinfeld & Sambhav Shrestha"
date: "06/24/2021"
output: html_document
---

***
## Introduction
We decided to replicate the data from the [article](https://github.com/msr-ds3/coursework/blob/master/week4/ft_police_complaints.pdf) on complaints filed against police officers in NYC, Chicago, and Philadelphia. All the data were taken from the links inside the article and wrangled and cleaned in 3 different R scripts for each cities as shown below.All the scripts for data wrangling can be found in the github. After, we plotted each graph for 3 cities and compared with the article.

```{r}
source('data_wrangling/02_clean_nyc_data.R')
source('data_wrangling/05_clean_chicago_data.R')
source('data_wrangling/08_clean_philly_data.R')
```



### New York City
```{r NYC graph}

nyc_incidents %>%
  ggplot(aes(x = decile_rank, y = num_complaints_in_this_decile / sum(num_complaints_in_this_decile))) +
  geom_bar(stat = "identity", aes(fill = decile_rank == 10)) +
  theme(legend.position = "none") +
  scale_x_continuous("Decile Rank", breaks = pretty_breaks(10)) +
  scale_y_continuous("% of Complaints", labels = percent) +
  labs(title = "NYC Complaints")

```


### Chicago 
```{r Chigago graph}
# plot the output
chicago_incidents %>%
  ggplot(aes(x = decile_rank, y = total_complaints/sum(total_complaints))) +
  geom_bar(stat = 'identity', aes(fill = decile_rank == 10)) +
  scale_y_continuous("% of complaints", labels = percent) +
  scale_x_continuous("Decile Rank", breaks = 1:10) +
  theme(legend.position = "none") +
  labs(title = "Chicago Complaints")
```



### Philadelphia 
```{r Philly graph}

philly_incidents %>%
  ggplot(aes(x = decile_rank, y = num_complaints_in_this_decile / sum(num_complaints_in_this_decile))) +
  geom_bar(stat = "identity", aes(color = decile_rank == 10, fill = decile_rank == 10)) +
  theme(legend.position = "none") +
  scale_x_continuous("Decile Rank", breaks = pretty_breaks(10)) +
  scale_y_continuous("% of Complaints", labels = percent, limits = c(0, 0.3)) +
  labs(title = "Philadelphia Complaints")

```

## Conclusion
The graphs obtained from NYC and Philadelphia matched perfectly whereas the Chicago graph missed the mark by 3% error. As presented in the article, more than 30% of complaints seem to come from police officers in the top 10% group in each cities.
















